---
title: Weathering
author: J. Hagedorn
date: '2018-06-05'
slug: weathering
categories: []
tags: ["weather"]
---

As a way to potentially incorporate weather data into poems and as a way to better get to know the forest and meadows by my house, I wanted to find a way to pull in historical weather information for a specific set of geographical coordinates.

I'm using the [`darksky`](https://github.com/hrbrmstr/darksky) package in R:

```{r libs, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(darksky); library(tidyverse);library(plotly)
```

First, I'll set my API key, available by [signing up for a DarkSky developer account](https://darksky.net/dev/register). You can set it for your own local environment using `Sys.setenv(DARKSKY_API_KEY = rstudioapi::askForPassword("Paste DarkSky API Key:"))`.  You can then test to ensure it's saved by running `Sys.getenv("DARKSKY_API_KEY")`.

Then, grab the last year's worth of hourly weather data for the lat/lon coordinates of interest:

```{r eval=FALSE, message=FALSE, warning=FALSE, include=TRUE, paged.print=FALSE}
blandford <-
seq(Sys.Date()-365, Sys.Date(), "1 day") %>% 
  map(~get_forecast_for(42.996041, -85.738570, .x)) %>% 
  map_df("hourly") 

write_csv(blandford,"../../static/data/blandford.csv")
```

In order to avoid calling the API multiple times, the above code chunk is not run and the data set is saved locally as a `.csv` file and loaded in the following script:

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
blandford <- read_csv("../../static/data/blandford.csv")
```

Below is an interactive plot of hourly temperature for an entire year, colored by % cloud cover.

```{r line, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
blandford %>%
  droplevels() %>%
  plot_ly(
    x = ~time,
    y = ~temperature,
    color = ~cloudCover,
    colors = colorRamp(c("#4da5ff","#A0A0A0"))
  ) %>%
  add_lines(
    alpha = 0.5
  ) %>%
  layout(
    xaxis = list(
      range = c(min(blandford$time), max(blandford$time))
    )
  )
```

