---
title: Starplots
author: J. Hagedorn
date: '2018-09-30'
slug: starplots
categories: []
tags:
  - astronomy
meta_img: /images/image.jpg
---

http://www.klima-luft.de/steinicke/ngcic/Expl_Hist_WH.htm

https://www.packtpub.com/mapt/book/web_development/9781787281776/32/ch11lvl1sec0048/creating-an-interactive-star-map



```{r}
library(tidyverse)

observed_stars <- 
  readxl::read_excel("../../static/data/Historic_WH.xls") %>%
  rename_all(funs(str_replace_all)," ","_") %>%
  rename(
    class = Class,
    catalogue = Cat,
    herschel_obj_n = H,
    sweep = Sw,
    year = Y,
    month = M,
    day = D,
    aperture = Ap,
    focus = Fo,
    observing_site = Loc,
    n_observations = Obs,
    catalogue_id = N,
    ngc_ic = NI,
    slough_catalog_id = h,
    general_catalog_id = GC
  )

```


```{r}
library(plotly)

observed_stars %>%
  plot_ly(
    x = ~RH,
    y = ~RM,
    z = ~RS
  ) 

```

Column  Explanation
Class 	Herschel designation: class (I...VIII) and running number in each class
Cat 	Herschel-catalogue (1, 2 or 3) which contains the object
H 	Herschel object number (1...2500)
Sw 	Number of discovery sweep
Y, M, D 	Discovery date as given in Herschel's published catalogues. Note: In the Historic NGC there can be given an earlier dat for some cases, based on his manuscripts.
Ap 	Aperture of discovery telescope (inch)
Foc 	Focus: N = Newtonian, F = front view
Loc 	Observing site: D = Datchet, C = Clay Hall (Old Winsor), S = Slough
Obs 	Number of observations (in different sweeps)
Identity 	Identity with other class object(s)
N 	Catalogue identifier: N = NGC, I = IC
NI 	NGC-/IC-number
h 	No. in John Herschel's Slough and Cape Catalogues
GC 	No. in John Herschel's General Catalogue
Discoverer 	Name of discoverer (and date if not W. Herschel)
Observer
	Independent (later) discoverer
RH ... DS
	Rect. & Decl. for 2000.0 (from Revised NGC/IC)
Vmag
	Visual magnitude (from Revised NGC/IC)
Con
	Constellation (from Revised NGC/IC)
Type
	Object type (from Revised NGC/IC)
S
	Object status: 1 = galaxy, 2 = galactic nebula, 3 = planetary nebula, 4 = open cluster, 5 = globular cluster, 6 = part of nebula, 7 = NGC dublicate, 9 = star(s), 10 = not found
ID1, ID2, ID3 	Messier number; double identifications in NGC or IC
