---
title: What will you say next?
author: J. Hagedorn
date: '2019-01-26'
slug: what-will-you-say-next
tags:
  - prosody
meta_img: /images/image.jpg
---

```{r libs, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
knitr::opts_chunk$set(
  echo=FALSE,message=FALSE,warning=FALSE,fig.width=6,fig.asp=0.618
)
library(tidyverse); library(gutenbergr); library(DT); library(sunburstR)
library(showtext)
font_add_google("Merriweather","merriweather")
```

### Caveat: 'We murder to dissect'

Voice is both a center, pouring from a single mouth, and also an environment: surrounding, echoing, reverberating, jarring, saturating. Walter J. Ong writes that, immersed in our hearing, we "gather sound simultaneously from every direction at once; I am at the center of my auditory world, which envelopes me, establishing me at a kind of core of sensation and existence..."^[from Ong, Walter J. *Orality and Literacy: The Technologizing of the Word*. Routledge, 2013.]

Across his essays, Ong plumbs the turn toward written language, cautioning that writing turns speech into a static, constrained and coffined thing.  Even at a remove of thousands of years from that schism, I can commiserate.  At times, feeling for a phrase, I have summoned sounds and nonsense words merely to detach words from the written characters to which they have been twined.  So it's strange that the honor he imparts to the voice echoed in my mind when I sat down to tinker with parsing patterns of poetic sentences.

Though I'm a bit biased, it seems that of any written speech, poetry especially bears investigation.  Ong opines that "[Rhythmically balanced, formulaic expressions] can be found occasionally in print but in oral cultures they...are incessant. They form the substance of thought itself."  Written poems are a late remnant of this music.

### Hopeful lepidopterists

Written language is everywhere, making it especially poignant to acknowledge what we have lost in branching away from oral culture.  Yet there is much to love in it.  One of the things that written language can give us, with its static wings, is a preserved body to stare into: beyond the expiration of air and the inter-personal awkwardness of staring.

Just as Ramon y Cahal segmented and sketched neuronal cells, yielding images which were both accurate and lovely, we might sketch out the patterns of utterance: not to be useful or prove some theory, but just to observe the unintentional patterns that a voice accumulates across  its expeditions.

We might hold out hope that words, even those committed to the page, continue to hold the sparks of personhood. As Coleridge proposes: "...if words are not THINGS, they are LIVING POWERS, by which the things of most importance to mankind are actuated, combined, and humanized."^[Coleridge, S.T. *Aids to Reflection*, ed. John Beer (1993), p. 10] 

### Seeing the hinges of speech

Nothing so ambitious is going on here, though.  Just a sketch of the patterns lying latent in the collected works of some beloved companions.  Of course by doing this, I'm probably proving Ong's point that "An oral culture simply does not deal in such items as geometrical figures, abstract categorization, formally logical reasoning processes, definitions, or even comprehensive descriptions, or articulated self-analysis, all of which derive not simply from thought itself but from text-formed thought."

The sunburst plot (below) resembles a flower's corolla, a liturgical fan, a geometric sun-flare.  Its petals are the unfolding sequential patterns of parts of speech^[The part-of-speech tagging in `cleanNLP` uses the [universal POS tags](https://universaldependencies.org/u/pos/index.html).  More detailed [Penn Treebank Project codes](https://www.ling.upenn.edu/courses/Fall_2003/ling001/penn_treebank_pos.html) are also available.]: those which follow each turn of a phrase, as signalled by a punctuation mark.^[We could also break up the pattern using sentences, line breaks or stanzas, but that will have to wait for another exploration.]  In the curled pause of a comma, at the hinge of a sentence, how does one coax the next thought?  Piecing together the collected utterances of a poet as a compound mouth, maybe we can catch a new angle of insight into what patterns accompany the summoning of a phrase.

Here are all of the phrases from Milton's *Paradise Lost*.  Starting from the blank hole at the center and proceeding outward, one can track the percent of phrases which fall into each part-of-speech sequence:^[I recognize that there are few people of either the literary or analytic bent who will likely want to read this, but for anyone interested in the code used to generate these visuals, see [the GitHub repo](https://github.com/j-hagedorn/journals_of/blob/master/content/projects/2019-01-26-parts-of-speech-patterns.Rmd)]

```{r annotate_poets, eval=FALSE}
library(cleanNLP)
cnlp_init_udpipe()
corpus <- feather::read_feather("../../static/data/corpus.feather")

milton <-
  corpus %>%
  filter(title == "Paradise Lost") %>%
  select(text) %>%
  # Remove titles and stanza breaks (all caps)
  filter(text != toupper(text)) %>%
  cnlp_annotate(.$text, as_strings = TRUE) %>%
  cnlp_get_token()

keats <-
  corpus %>%
  filter(author == "Keats, John") %>%
  select(text) %>%
  # Remove titles and stanza breaks (all caps)
  filter(text != toupper(text)) %>%
  cnlp_annotate(.$text, as_strings = TRUE) %>%
  cnlp_get_token()

feather::write_feather(milton,"../../static/data/milton.feather")
feather::write_feather(keats,"../../static/data/keats.feather")
```

```{r get_annotations}
milton <- feather::read_feather("../../static/data/milton.feather")
keats <- feather::read_feather("../../static/data/keats.feather")
```

```{r fig.cap= "Part-of-speech sequences from Milton's Paradise Lost"}
df <-
  milton %>%
  # Remove apostrophes
  filter(str_detect(word,"'") == F) %>%
  mutate(
    word = str_replace_all(word,"'",""),
    punct = cumsum(upos == "PUNCT")
  ) %>%
  filter(upos != "PUNCT") %>%
  group_by(punct) %>%
  summarize(
    path = paste(tolower(upos), collapse = "-"),
    phrase = paste(word, collapse = " "),
    path_size = n()
  ) %>%
  # Only include phrases with more than one word
  filter(path_size > 1) %>%
  group_by(path) %>%
  mutate(score = n()) %>%
  arrange(score) %>% 
  # Only include patterns which occur more than once
  filter(score > 1) %>%
  ungroup() 

df %>% 
  select(path,score) %>% 
  distinct() %>% 
  arrange(path) %>% 
  sunburst()

# sund2b(df)
```

In Milton's *Paradise Lost* the most common patterns, if we move outward from the largest segments at the center, include: 

- `cconj-noun-adp`, from phrases such as "`r df %>% filter(str_detect(path,"^cconj-noun-adp")) %>% head(10) %>% .$phrase %>% paste(.,collapse = '", "')`"
- `verb-det-noun`, found in phrases like "`r df %>% filter(str_detect(path,"^verb-det-noun")) %>% head(10) %>% .$phrase %>% paste(.,collapse = '", "')`"

What if we look at a different poet's work?  Will it look obviously different, or does the form necessitated by the language itself (the ligaments of its connections) overwhelm any patterns peculiar to a single person?^[Is this an instance of the models we use (e.g. language) speaking louder than our individual voices, or a story about the inadequacy of models to describe the utterance of an individual riding its rules?  Are those stories the same one, with different narrators? ]  Here are the collected poems of Keats:

```{r fig.cap = "Part-of-speech sequences from Keats' selected poems"}
df <-
  keats %>%
  # Remove apostrophes
  filter(str_detect(word,"'") == F) %>%
  mutate(
    word = str_replace_all(word,"'",""),
    punct = cumsum(upos == "PUNCT")
  ) %>%
  filter(upos != "PUNCT") %>%
  group_by(punct) %>%
  summarize(
    path = paste(tolower(upos), collapse = "-"),
    phrase = paste(word, collapse = " "),
    path_size = n()
  ) %>%
  # Only include phrases with more than one word
  filter(path_size > 1) %>%
  group_by(path) %>%
  mutate(score = n()) %>%
  arrange(score) %>% 
  # Only include patterns which occur more than once
  filter(score > 1) %>%
  ungroup() 

df %>% 
  select(path,score) %>% 
  distinct() %>% 
  arrange(path) %>% 
  sunburst()

# legend_items <- unique(unlist(strsplit(df$path, '-')))
# cols <- viridis::viridis_pal(option = "viridis")(length(legend_items))

# df %>% sunburst(
#       colors = list(
#       range=cols,
#       domain=legend_items
#     )
#   )

```

In Keats' poems the most common patterns, if we move outward from the largest segments at the center, include: 

- `cconj-adp-det`, from phrases such as "`r df %>% filter(str_detect(path,"^cconj-adp-det")) %>% head(10) %>% .$phrase %>% paste(.,collapse = '", "')`"
- `adp-det-adj-noun`, found in phrases like "`r df %>% filter(str_detect(path,"^adp-det-adj-noun")) %>% head(10) %>% .$phrase %>% paste(.,collapse = '", "')`"

I'm tempted to find in these phrases something typically "Keatsian", and in the previous ones something "Miltonic", but minds are quick to latch onto a good story in the absence of evidence.  Nor do I think that there's anything particular like "evidence" for the phenomenon I'm interested in: personhood, and its manifestation in language.  That said, I do love patterns.  Reductive and lovely as they are, when they are made visible...

To get an inkling of whether there are patterns that are peculiar to one poet more than another, I did a little sequential association rule mining.^[This analysis uses the association rule mining package referenced in Hahsler M, et al. (2005a). *"arules – A Computational Environment for Mining Association Rules and Frequent Item Sets."* Journal of Statistical Software, 14(15), 1–25.  It also uses `arulessequences`, an add-on for `arules` to handle and mine frequent sequences.Provides interfaces to the C++ implementation of cSPADE byMohammed J. Zaki.]  With this approach, pattern sequences are scored using the following characteristics for each body of work (Milton & Keats):

- *How common?*: Often referred to as the `support` for an associated bundle, this is the percentage of all times that this bundle of services occurred somewhere in a person's service history.
- *Strength*: This is the strength of the association between the bundle of services. Specifically, it tells us what percent of the time that the service on the right hand side (or column) is provided if the service(s) on the left side are provided.
- *How much more likely?*: Seeing the patterns on the left hand side makes it this many times more likely that we'll see the pattern on the right occur next. More specifically, it is the probability that we actually see the patterns in sequence compared to the probability that we would see them together if they were unrelated.  Technically, it's called `lift`.

In the plot below, each dot stands for a rule/pattern which is found in both poets' work, using the `lift` measure described just above. The colors are based on difference, so that yellow means "this pattern is pretty similar between both" and purple means "this pattern differs most between them".  Some of those sequential part-of-speech patterns are shown for reference.

```{r fun_seq_rules}
get_seq_rules <- function(df,support = 0.5,confidence = 0.5){
  
  library(tidyverse); library(arules); library(arulesSequences)
  
  seq_df <-
    df %>%
    # Remove apostrophes
    filter(str_detect(word,"'") == F) %>%
    mutate(
      word = str_replace_all(word,"'",""),
      sequenceID = cumsum(pos == ".") + 1
    ) %>%
    group_by(sequenceID) %>%
    mutate(eventID = cumsum(upos == "PUNCT") + 1) %>%
    filter(upos != "PUNCT") %>%
    select(sequenceID,eventID,upos) %>%
    group_by(sequenceID,eventID) %>%
    summarize(
      size = n(),
      items = paste(tolower(upos), collapse = ";")
    ) %>%
    group_by(sequenceID) %>%
    mutate(eventID = row_number()) %>%
    distinct()
  
  # Convert to transaction matrix data type
  write.table(
    seq_df,file.path(tempdir(),"seq_df.txt"),
    sep=";",row.names=F,col.names=F,quote = F
  )
  
  baskets <- 
    read_baskets(
      file.path(tempdir(),"seq_df.txt"), sep = ";", 
      info = c("sequenceID","eventID","size")
    )
  
  # Get frequent sequences and corresponding support values
  freq_seq <- 
    cspade(
      baskets, 
      parameter = list(support = support), 
      control = list(verbose = TRUE)
    ) 
  
  rule_df <- 
    ruleInduction(
      freq_seq,
      confidence = confidence,
      control = list(verbose = T)
    ) %>%
    as("data.frame")
  
  return(rule_df)
  
}
```

```{r get_rules,include=FALSE}
seq_rules_milton <- get_seq_rules(milton,0.5,0.5)
seq_rules_keats <- get_seq_rules(keats,0.3,0.5)

seq_rules_df <- 
  seq_rules_milton %>%
  full_join(seq_rules_keats, by = "rule") %>%
  rename_at(vars(ends_with(".x")),list(~str_replace(.,".x$","_milton"))) %>%
  rename_at(vars(ends_with(".y")),list(~str_replace(.,".y$","_keats")))

rm(seq_rules_keats); rm(seq_rules_milton)
```

```{r rule_scatter,fig.showtext=TRUE,dpi=300,fig.retina=1}
# Note: sizing needs to be larger for high DPI

showtext_auto()

seq_rules_df %>%
  filter(complete.cases(.)) %>%
  mutate(
    difference = lift_milton - lift_keats,
    some_labels = if_else(
      percent_rank(difference) > 0.993,
      rule,NA_character_
    )
  ) %>%
  ggplot(
    aes(
      x = lift_milton, y = lift_keats, 
      color = difference, label = some_labels
    )
  ) +
  geom_point(alpha = 0.5) +
  ggrepel::geom_label_repel(size = 6) +
  scale_color_viridis_c(direction = -1) +
  labs(
    title = "Comparative part-of-speech sequencing",
    subtitle = "If pattern X occurs, how many times more likely is pattern Y to come next...",
    x = "...in Milton's 'Paradise Lost'?",
    y = "...in Keats' collected poems?"
  ) +
  theme_minimal() +
  theme(text=element_text(size = 25,family="merriweather"))


```

Not much "aha!" for this introductory exploration.  We could pour in more poems, and include comparisons to instruction manuals, Audobon guides, religious texts, etc.  And with all this, any pattern found would still likely have only a tinny resonance in comparison to the poems themselves.  In addition to the aesthetic sublime, though, there is a mathematical one: one not necessarily tuned to the human ear or heart.  

In the afterword to his collection *Museum*, Jorge Luis Borges speculates that all of the wanderings which a person makes eventually trace a recognizable shape:

> "A man sets out to draw the world.  As the years go by, he peoples a space with images of provinces, kingdoms, mountains, bays, ships, islands, fishes, rooms, instruments, stars, horses, and individuals.  A short time before he dies, he discovers that that patient labyrinth of lines traces the lineaments of his own face."^[Borges, J.L. *Collected Fictions*. trans. Andrew Hurley. New York: Penguin, 1998.]

A footprint in the dust is not equal to a foot, certainly. And yet the traces that we leave hold some remnant of us, beyond our intentions and our desire to trace any particular legacy.  And if the momentary accident of a single phrase can delight with a striking analogy, perhaps the accidental or unintentional patterns of phrasing over the course of one’s life may also spark some slight insight.
